\propertitle

### 2019-11-17 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-17T00:27:31+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

\newcommand*{\sh}{\mathrm{\varEta}}
%%% Custom macros end @@@

@@

We measure the difference $\ydi(M'',M')$ in informational sufficiency
between a set of moments, say $M''\defd\set{1,\dotsc,m''}$, and another, say
$M'\defd \set{1,\dotsc,m'}$, as follows:
\begin{enumerate}[label=(\roman*)]
\item from each maximum-entropy distributions $\px(\yA \| M)$ for the full
  network, built from each set of constraints $M=M',M''$, calculate the
  marginal distribution for the sample:
\begin{equation}
  \label{eq:sample_marginal}
  \pxx(\ya \| M) = \sum_{\yA}\yG_{\ya \yA}\px(\yA \| M),
  \qquad M=M',M'';
\end{equation}
\item calculate the relative entropies of the measured frequency
  distribution $\yf$ with respect to each sample marginal, and multiply
  them by the number of time bins $T$:
  \begin{equation}
    \label{eq:relentropy_T}
    T\,\sh[\yf; \pxx(\ya \| M)] \defd  T  \sum_{\ya} \yff_{\ya} \ln\frac{\yff_{\ya}}{
      \pxx(\ya \| M)},
    \qquad M=M',M'';    
  \end{equation}
\item take the difference: 
\begin{multline}
  \label{eq:measure_suffic}
  \ydi(M'',M') \defd
  T \,\sh[\yf; \pxx(\ya \| M')]
  -
  T \,\sh[\yf; \pxx(\ya \| M'')]
  \equiv{}\\
T  \sum_{\ya} \yff_{\ya} \ln\frac{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M'')}{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M')}.
\end{multline}
\end{enumerate}
The measure $\ydi(M'',M')$ so defined is positive if $M''$ is \enquote{more
  informationally sufficient} than $M'$.

Why is this a natural measure? Because, if we assign equal pre-data
probabilities to the two hypotheses $M''$ and $M'$, then the exponential of
$\ydi(M'',M')$ is approximately equal to the ratio of their probabilities
conditional on the data $\yf$ -- their Bayes factor:
\begin{equation}
  \label{eq:bayes_factor}
  \ydi(M'',M') \approx
  \ln\bigl[\pf(M'' \| \yf)/
  \pf(M' \| \yf)\bigr].
\end{equation}
So the exponential of $\ydi(M'',M')$ tells us how much more probable $M''$
is than $M'$, in view of the data $\yf$. We prove this in appendix**

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/comparison_full.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/trunct_comparison_marginals.pdf}%
\caption{***}
\label{fig:comparisons_marginals}
\end{figure}
% comparison_ME_sample_full_levels.nb
%
% relentropies of sample2mom, sample4mom, sample5mom, marg2mom, marg4mom, marg5mom
% -322.72,-42.8005,-39.4917,-85.7626,-4.23725,-4.20059 nats =
% -1401.56,-185.88,-171.51,-372.462,-18.4021,-18.2429 dB
%
% difference in relentropies between 4 mom and 2 mom:
% sample level: 279.92 nats = 1215.68 dB
% marg level: 81.525 nats =  354.06 dB
% difference in relentropies between 5 mom and 4 mom:
% sample level: 3.30882 nats = 14.37 dB
% marg level: 0.0366623 nats = 0.159223 dB
% correspond to prob ratios of
% sample: 3.69524\times 10^{121}, 27.3529
% marg: 2.5469\times 10^{35}, 1.03734
% 
% comparison_ME_sample_full_levels.nb
%

@@

The second point is that our degrees of belief about the frequency
distribution for the full network depend not only on the measured data in
the sample, but also on our pre-data beliefs $\yI$ about the distribution.
Which assumptions lead to the maximum-entropy result? This distribution
appears when our initial belief about the possible frequency distributions
$\yF$ is quantified by an entropic prior
\citep{neumann2007,rodriguez1991,skilling1998,catichaetal2004,portamana2017}:
\begin{equation}
  \label{eq:entropic_prior}
  \pf(\yF \| \yI) \propto  \exp[-\yL\;\re(\yF;\yR)]
  \approx \binom{L}{L\yFF_{0}, \dotsc, L\yFF_{N}}\;
  \prod_{\yA}{\yRR_{\yA}}^{L\yFF_{\yA}}
\end{equation}
where $\re(\yF;\yR)\defd \sum_{\yA}\yFF_{\yA}\ln(\yFF_{\yA}/\yRR_{\yA})$ is
the relative entropy or discrimination information
\citep{kullback1987,jaynes1963,hobson1969,hobsonetal1973}, $\yR$ is the
reference distribution, and $\yL$ a positive parameter. The approximate
equality (obtained through Stirling's approximation), where the large
parentheses denote a multinomial coefficient, shows that this prior belief
is proportional to the number of ways in which the distribution $\yF$ can
be realized in $L$ time bins. The parameter $L$ roughly quantifies how many
time bins our data set must have to affect our initial belief. The
maximum-entropy approximation is valid when $L$ is large, but small
compared to the sharpness of the constraints on $\yF$; in our case this
means $\yL \approx 10$, give or take an order of magnitude.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-17T15:08:12+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

\newcommand*{\sh}{\mathrm{\varEta}}
\newcommand*{\nat}{\textrm{nat}}
\newcommand*{\hart}{\textrm{Hart}}
%%% Custom macros end @@@

@@

We measure the difference $\ydi(M'',M')$ in informational sufficiency
between a set of moments, say $M''\defd\set{1,\dotsc,m''}$, and another, say
$M'\defd \set{1,\dotsc,m'}$, as follows:
\begin{enumerate}[label=(\roman*)]
\item from each maximum-entropy distributions $\px(\yA \| M)$ for the full
  network, built from each set of constraints $M=M',M''$, calculate the
  marginal distribution for the sample:
\begin{equation}
  \label{eq:sample_marginal}
  \pxx(\ya \| M) = \sum_{\yA}\yG_{\ya \yA}\px(\yA \| M),
  \qquad M=M',M'';
\end{equation}
\item calculate the relative entropies of the measured frequency
  distribution $\yf$ with respect to each sample marginal, and multiply
  them by the number of time bins $T$:
  \begin{equation}
    \label{eq:relentropy_T}
    T\,\sh[\yf; \pxx(\ya \| M)] \defd  T  \sum_{\ya} \yff_{\ya} \log\frac{\yff_{\ya}}{
      \pxx(\ya \| M)},
    \qquad M=M',M'';    
  \end{equation}
\item take the difference: 
\begin{multline}
  \label{eq:measure_suffic}
  \ydi(M'',M') \defd
  T \,\sh[\yf; \pxx(\ya \| M')]
  -
  T \,\sh[\yf; \pxx(\ya \| M'')]
  \equiv{}\\
T  \sum_{\ya} \yff_{\ya} \log\frac{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M'')}{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M')}.
\end{multline}
\end{enumerate}
The measure $\ydi(M'',M')$ so defined is positive if $M''$ is \enquote{more
  informationally sufficient} than $M'$.

Why is this a natural measure? Because $\ydi(M'',M')$ is equal to the
log-ratio of the likelihoods of the hypotheses $M''$ and $M'$ in view of
the data $\yf$:
\begin{equation}
  \label{eq:bayes_factor}
  \ydi(M'',M') =
  \log\bigl[\pf(\yf \| M'')/
  \pf(\yf \| M')\bigr].
\end{equation}
This is called their \emph{relative weight of evidence}, the logarithm of
their \emph{relative Bayes factor}
\citep[\chap~6]{good1950}{good1975,good1981,good1985,good1983}[\sect~1.4]{osteyeeetal1974}{mackay1992,kassetal1995}[see
also][p.~421]{jeffreys1936}[\chaps~V, VI, A]{jeffreys1939_r1983}. We prove
this in appendix**. So the exponential (same base) of $\ydi(M'',M')$ tells
us how much more probable the data $\yf$ are, in view of $M''$ vs $M'$. If
we also introduce the pre-data probabilities of the two hypotheses we
obtain their probabilities conditional on the data.

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/comparison_full.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/trunct_comparison_marginals.pdf}%
\caption{***}
\label{fig:comparisons_marginals}
\end{figure}
% comparison_ME_sample_full_levels.nb
%
% relentropies of sample2mom, sample4mom, sample5mom, marg2mom, marg4mom, marg5mom
% -322.72,-42.8005,-39.4917,-85.7626,-4.23725,-4.20059 nats =
% -1401.56,-185.88,-171.51,-372.462,-18.4021,-18.2429 dB
%
% difference in relentropies between 4 mom and 2 mom:
% sample level: 279.92 nats = 1215.68 dB
% marg level: 81.525 nats =  354.06 dB
% difference in relentropies between 5 mom and 4 mom:
% sample level: 3.30882 nats = 14.37 dB
% marg level: 0.0366623 nats = 0.159223 dB
% correspond to prob ratios of
% sample: 3.69524\times 10^{121}, 27.3529
% marg: 2.5469\times 10^{35}, 1.03734
% 
% comparison N vs n:
% 2 mom: 236.958 nats = 1029.09 dB, 8.11758\times 10^{102}
% 4 mom: 38.5632 nats = 167.478 dB, 5.59496*10^16
% 5 mom: 35.2911 nats = 153.267 dB, 2.12185*10^15
%
% comparison_ME_sample_full_levels.nb
%

Consider the three sets of constraints $M_{2}$, $M_{4}$, $M_{5}$ consisting
in the first two, four, five moments. For a full network of size
$N=10\,000$, we obtain the following differences:
\begin{equation}
  \label{eq:diff_suff_N}
  \begin{split}
  \ydi(M_{4},M_{2}) &= 81\;\nat = 35\;\hart,
\\    \ydi(M_{5},M_{4}) &= 0.037\;\nat = 0.016\;\hart,
  \end{split}
\end{equation}
where the Hartley ($\hart$) denotes base-10 logarithms
\citep[\sect~C.4]{iso2009}. In words, the measured frequencies of the
sample activity are 35 orders of magnitude more probable under sufficiency
of the first four moments are sufficient, than under sufficiency of the
first two only; but they are about equally probable ($10^{0.016} = 1.04$)
whether we consider the first five moments to be sufficient or just the
first four.

Compare the results above with those obtained by applying the method at the
sample level only:
\begin{equation}
  \label{eq:diff_suff_n}
  \begin{split}
  \ydi_{\text{sample level}}(M_{4},M_{2}) &= 280\;\nat = 1\,220\;\hart,
\\    \ydi_{\text{sample level}}(M_{5},M_{4}) &= 3.3\;\nat = 1.4\;\hart,
\end{split}
\end{equation}
that is, the data are 1\,220 orders of magnitude more probable conditional
on four moments as opposed to two, and $10^{1.4}=25$ times more probable
conditional on five moments as opposed to five.

The application at the full-network level therefore leads to different
conclusions about the relative importance of the moments from the
application at the sample level. It is also possible to use the
measure~\eqref{eq:measure_suffic} and its probabilistic
meaninig~\eqref{eq:bayes_factor} to compare the probability of the data
$\yf$ conditional on the hypothesis $M_{5, N}$ of 5-moment sufficiency at
the full-network level, $N=10\,000$, versus the hypothesis $M_{5, n}$ of
5-moment sufficiency at the sample level, $N=n=65$. We obtain
\begin{equation}
  \label{eq:diff_suff_n}
  \ydi(M_{5, N}, M_{5, n}) = 35\;\nat = 15\;\hart,
\end{equation}
that is, the data are 15 orders of magnitude more likely under the first
hypothesis than under the second.

@@

The second point is that our degrees of belief about the frequency
distribution for the full network depend not only on the measured data in
the sample, but also on our pre-data beliefs $\yI$ about the distribution.
Which assumptions lead to the maximum-entropy result? This distribution
appears when our initial belief about the possible frequency distributions
$\yF$ is quantified by an entropic prior
\citep{neumann2007,rodriguez1991,skilling1998,catichaetal2004,portamana2017}:
\begin{equation}
  \label{eq:entropic_prior}
  \pf(\yF \| \yI) \propto  \exp[-\yL\;\re(\yF;\yR)]
  \approx \binom{L}{L\yFF_{0}, \dotsc, L\yFF_{N}}\;
  \prod_{\yA}{\yRR_{\yA}}^{L\yFF_{\yA}}
\end{equation}
where $\re(\yF;\yR)\defd \sum_{\yA}\yFF_{\yA}\log(\yFF_{\yA}/\yRR_{\yA})$ is
the relative entropy or discrimination information
\citep{kullback1987,jaynes1963,hobson1969,hobsonetal1973}, $\yR$ is the
reference distribution, and $\yL$ a positive parameter. The approximate
equality (obtained through Stirling's approximation), where the large
parentheses denote a multinomial coefficient, shows that this prior belief
is proportional to the number of ways in which the distribution $\yF$ can
be realized in $L$ time bins. The parameter $L$ roughly quantifies how many
time bins our data set must have to affect our initial belief. The
maximum-entropy approximation is valid when $L$ is large, but small
compared to the sharpness of the constraints on $\yF$; in our case this
means $\yL \approx 10$, give or take an order of magnitude.
